\documentclass{article}
\usepackage[submission]{colm2026_conference}

\usepackage{microtype}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{subcaption}
\newcommand{\nicefrac}[2]{#1/#2}
\usepackage{caption}
\usepackage{pifont}
\usepackage{lineno}

\definecolor{darkblue}{rgb}{0, 0, 0.5}
\hypersetup{colorlinks=true, citecolor=darkblue, linkcolor=darkblue, urlcolor=darkblue}

\captionsetup{font=small, labelfont=bf}

\title{SenseMath: A Controlled Benchmark for Shortcut-Sensitive Numerical Reasoning in Large Language Models}

\author{
  Anonymous Submission
}

\begin{document}

\ifcolmsubmission
\linenumbers
\fi

\maketitle

\begin{abstract}
\input{sections/0_abstract}
\end{abstract}

\input{sections/1_introduction}
\input{sections/2_related_work}
\input{sections/3_benchmark}
\input{sections/4_experiments}
\input{sections/5_conclusion}

\bibliography{references}
\bibliographystyle{colm2026_conference}

\appendix

\section{Example Items}
\label{sec:appendix_examples}

Table~\ref{tab:examples} shows example \textsc{SenseMath} items at $d{=}2$ for four categories.

\input{figures/TABLE_examples.tex}

\section{Prompt Templates}
\label{sec:appendix_prompts}

\textbf{CoT:} ``Please solve the following multiple-choice problem. Show your step-by-step reasoning, then provide your final answer as a single capital letter (A, B, C, or D) inside \textbackslash boxed\{\}.''

\textbf{NS:} ``Please solve the following multiple-choice problem using only easy calculations. Rely on your mathematical intuition, number sense, and estimation. Show brief reasoning, then provide your final answer as a single capital letter (A, B, C, or D) inside \textbackslash boxed\{\}.''

\textbf{Strict:} ``Please solve the following multiple-choice problem by performing precise, step-by-step arithmetic. Do NOT use estimation, rounding, or shortcuts. Show every computation digit by digit, then provide your final answer as a single capital letter (A, B, C, or D) inside \textbackslash boxed\{\}.''

\section{Task Specifications}
\label{sec:appendix_tasks}

\paragraph{J1 prompt:} ``Consider this math problem. Could you solve it significantly faster using mental math or a clever observation, compared to computing it step by step? Answer YES or NO, then explain in one sentence why.''

\paragraph{J2 prompt:} ``Classify the solution strategy as SHORTCUT or COMPUTATION. SHORTCUT: uses estimation, approximation, pattern recognition, structural insight, cancellation, or benchmark comparison. COMPUTATION: performs standard step-by-step arithmetic.''

\paragraph{G2:} Models receive a category description and one example strong/control pair, then generate a new pair in JSON format. Six deterministic code checks verify: (1) strong answer correctness, (2) control answer correctness, (3) shortcut existence in strong, (4) shortcut absence in control, (5) family matching, (6) novelty and digit scale.

\paragraph{Heuristic validation.} A deterministic shortcut solver applies category-specific heuristics without exact arithmetic. On strong items it achieves $\geq 70\%$; on control items approximately 25\%, confirming a 40pp separation.

\paragraph{Benchmark integrity.} All distractors share the last 50\% of digits (arithmetic) or fall within $\pm 0.1$ decimal (fractions). Correct option position is balanced across A/B/C/D.

\section{Strict Condition Results}
\label{sec:appendix_strict}

Table~\ref{tab:strict} presents Strict accuracy at $d{=}4$. High truncation for open-weight models (76--97\%) deflates their estimates.

\input{figures/TABLE_strict.tex}

\end{document}
